Inference in Hidden Markov Models with Explicit State Duration Distributions

### 2011-11-02 tex/introduction.tex
crunch crunch crunch (also a couple of typos, updated description of number of transitions graph)

The explicit state duration HMM (EDHMM) \cite{Rabiner89} was developed to allow explicit parameterization and direct inference of state duration distributions.  Approximate EDHMM estimation and inference can, in some cases, be performed using a modified forward-backward algorithm; specifically if either the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.   If the sequence is short then forward-backward can be run on a state representation that allows for all possible durations shorter than the observed sequence length.  If the sequence is long then forward-backward only remains computationally tractable if only transitions between ``allowed'' durations (those that lie within pre-specified allowable intervals) are considered.   If the true state durations lie outside those intervals then the resulting model estimates will be incorrect in the sense that the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals (which may poorly describe the true duration distributions).

Our contribution is the development of a procedure for EDHMM inference that does not require any hard pre-specification of duration intervals and, as a result, is no longer approximate.  The technique we use to do this 
is borrowed from sampling procedures developed for nonparametric  Bayesian HMM variants \cite{vanGael2008}.  Our key insight is simple: the machinery developed for doing inference in HMMs with a countable number of states is precisely the same as that which is needed for doing inference in an EDHMM with duration distributions over countable support.  So, while the EDHMM is a distinctly parametric model, the tools from nonparametric Bayesian inference can be applied such that black-box inference becomes possible and, in fact, efficient.

>>>>

The explicit state duration HMM (EDHMM) \cite{Rabiner89} was developed to allow explicit parameterization and direct inference of state duration distributions.  Approximate EDHMM estimation and inference can, in some cases, be performed using a modified forward-backward algorithm; specifically if the sequence is short or a tight ``allowable'' duration interval for each state is hard-coded a priori \cite{Yu2006}.   If the sequence is short then forward-backward can be run on a state representation that allows for all possible durations shorter than the observed sequence length.  If the sequence is long then forward-backward only remains computationally tractable if only transitions between durations that lie within pre-specified allowable intervals are considered.   If the true state durations lie outside those intervals then the resulting model estimates will be incorrect: the learned duration distributions will reflect only what is allowed given the pre-specified duration intervals (which may poorly describe the true duration distributions).

Our contribution is the development of a procedure for EDHMM inference that does not require any hard pre-specification of duration intervals and, as a result, is no longer approximate.  The technique we use to do this 
is borrowed from sampling procedures developed for nonparametric  Bayesian HMM variants \cite{vanGael2008}.  Our key insight is simple: the machinery developed for inference in HMMs with a countable number of states is precisely the same as that which is needed for doing inference in an EDHMM with duration distributions over countable support.  So, while the EDHMM is a distinctly parametric model, the tools from nonparametric Bayesian inference can be applied such that black-box inference becomes possible and, in fact, efficient.

### 2011-11-02 tex/results.tex
crunch crunch crunch (also a couple of typos, updated description of number of transitions graph)

@@

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_means.pdf}
        \label{fig:posterior_means}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates.pdf}
        \label{fig:posterior_rates}
    }
    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters for the data shown in Figure \ref{fig:experiment1_data}.}
    \label{fig:experiment1_results}
\end{figure}
\subsection{Synthetic Data}

The first experiment uses the 500 data points shown in Figure \ref{fig:experiment1_data} generated from an EDHMM with three states. To generate the data the duration distributions were chosen to be Poisson with rates $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$; each observation distribution was Gaussian with means of $\mu_1 = -3$, $\mu_2 = 0$, and $\mu_3 = 3$, each with a variance of 1. The transition distributions $A$ were set to
\begin{equation*}
\begin{bmatrix}
    0 & 0.3 & 0.7 \\ 0.6 & 0 & 0.4 \\ 0.3 & 0.7 & 0
\end{bmatrix}.  
\end{equation*}

For inference, broad, uninformative priors were chosen for the parameters of the duration and observation distributions. The parameters of the observation distribution for all states were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$. The rate parameters for all states were given $\mathrm{Gamma}(1, 10^{5})$ priors.

One thousand samples were collected from the EDHMM beam sampler after a burn-in of 500 samples. The learned posterior distribution of the state duration parameters and means of the observation distributions are shown in Figure \ref{fig:experiment1_results}.  The EDHMM achieves high accuracy in the estimated posterior distribution of the observation means, despite the overlap in observation distributions. The rate parameter distributions are reasonably estimated given the small number of observed segments.

A second experiment was performed using similar synthetic data to demonstrate the ability of the EDHMM to distinguish between states having differing duration distributions but the same observation distribution. Here the same model was used as above except data 
%was 
were
generated using $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$. The same priors were used as above, and again the sampler was allowed to run for 1000 samples with a burn-in of 500 samples.

Figure~\ref{fig:experiment2_results} shows that the sampler clearly separates the high state associated with $\mu_3$ from the other two states and clearly 
%identifies the fact that there are 
reveals the presence of
two  low states with differing duration distributions.   That it is possible for the EDHMM to recover two states with differing duration distributions requires a third state.  Figure~\ref{fig:exp_2_state} shows posterior samples that indicate that the model is mixing over ambiguities about states $0$ and $1$ as it should. %As shown in Figure \ref{fig:experiment2_results}, when the state transitions to and from the long-high state to the short-low state (e.g. $t\approx 100$ and $t\approx 180$) the transitions are captured well. When the system transitions between the two low states the segmentation is less accurate, due to the inherent lack of identifiability during these periods. Instead, the space of valid state sequences is explored, each sequence obeying the duration distribution.

@@

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_Y.pdf}
        \label{fig:exp_2_data}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_X.pdf}
        \label{fig:exp_2_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_means_exp3.pdf}
        \label{fig:posterior_means_3}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates_exp3.pdf}
        \label{fig:posterior_rates_3}
    }
    \caption{Results of the beam sampler applied to a system that generates states with identical observation distributions but differing durations. The observations are shown in a), and the true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler. Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$. Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
    \label{fig:experiment2_results}
\end{figure}

Finally, Figure \ref{fig:allowed} shows that the number of allowed transitions decays with the number of iterations, significantly speeding up the runtime of the later samples after the burn-in period.

\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in iterations on data from Figure \ref{MIKE}. Consider this in comparison to the $(KT)^2 = O(10^6)$ transitions that would need to be considered by standard forward backward without truncation, the only surely safe, truncatation-free alternative.}
\label{fig:allowed}
\end{figure}

>>>>

\subsection{Synthetic Data}

The first experiment uses the 500 data points (Figure \ref{fig:experiment1_data}) generated from a three state EDHMM. The duration distributions were Poisson with rates $\lambda_1 = 5$, $\lambda_2 = 15$, $\lambda_3 = 20$; each observation distribution was Gaussian with means of $\mu_1 = -3$, $\mu_2 = 0$, and $\mu_3 = 3$, each with a variance of 1. The transition distributions $A$ were set to
\begin{equation*}
\begin{bmatrix}
    0 & 0.3 & 0.7 \\ 0.6 & 0 & 0.4 \\ 0.3 & 0.7 & 0
\end{bmatrix}.  
\end{equation*}

@@

Broad, uninformative priors were chosen for the parameters of the duration and observation distributions. The the observation distribution parameters were given a normal-inverse-Wishart (N-IW) prior with parameters $\nu_0 = 2$, $\Lambda_0 = 1$, $\kappa=0.1$ and $\mu_0 = 0$. The rate parameters for all states were given $\mathrm{Gamma}(1, 10^{5})$ priors.

One thousand samples were collected from the EDHMM beam sampler after a burn-in of 500 samples. The learned posterior distribution of the state duration parameters and means of the observation distributions are shown in Figure \ref{fig:experiment1_results}.  The EDHMM achieves high accuracy in the estimated posterior distribution of the observation means, despite the overlap in observation distributions. The rate parameter distributions are reasonably estimated given the small number of observed segments. Figure \ref{fig:allowed} shows the mean number of transitions visited per time point over each iteration of the sampler.

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_means.pdf}
        \label{fig:posterior_means}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates.pdf}
        \label{fig:posterior_rates}
    }
    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters for the data shown in Figure \ref{fig:experiment1_data}.}
    \label{fig:experiment1_results}
\end{figure}

\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in iterations on data from Figure \ref{fig:experiment1_results}. Consider this in comparison to the $(KT)^2 = O(10^6)$ transitions that would need to be considered by standard forward backward without truncation, the only surely safe, truncation-free alternative.}
\label{fig:allowed}
\end{figure}

A second experiment was performed to demonstrate the ability of the EDHMM to distinguish between states having differing duration distributions but the same observation distribution. The same model and sampling procedure was used as above except here $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$.

Figure~\ref{fig:experiment2_results} shows that the sampler clearly separates the high state associated with $\mu_3$ from the other states and clearly 
%identifies the fact that there are 
reveals the presence of
two  low states with differing duration distributions.   
%That it is possible for the EDHMM to recover two states with differing duration distributions requires a third state.  
Figure~\ref{fig:exp_2_state} shows posterior samples that indicate that the model is mixing over ambiguities about states $0$ and $1$ as it should. %As shown in Figure \ref{fig:experiment2_results}, when the state transitions to and from the long-high state to the short-low state (e.g. $t\approx 100$ and $t\approx 180$) the transitions are captured well. When the system transitions between the two low states the segmentation is less accurate, due to the inherent lack of identifiability during these periods. Instead, the space of valid state sequences is explored, each sequence obeying the duration distribution.

@@

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_Y.pdf}
        \label{fig:exp_2_data}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_X.pdf}
        \label{fig:exp_2_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_means_exp3.pdf}
        \label{fig:posterior_means_3}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates_exp3.pdf}
        \label{fig:posterior_rates_3}
    }
    \caption{Beam sampler results applied to a system with identical observation distributions but differing durations. Observations are shown in a); true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler. Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$. Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
    \label{fig:experiment2_results}
\end{figure}

### 2011-11-02 tex/sampling.tex
crunch crunch crunch (also a couple of typos, updated description of number of transitions graph)

A recent contribution to inference 
in 
%is 
the infinite Hidden Markov Model (iHMM) \cite{Beal2002} suggests a way around truncation \cite{vanGael2008}.  The iHMM is an HMM with a countable number of states.  Computing the forward message for a forward filtering, backward sampler for the latent states in an iHMM also requires a sum over a countable number of elements.  
% which can be dealt with using the beam sampler \cite{vanGael2008}. Here, instead of a %potentially infinite duration-space cardinality, the iHMM faces a potentially infinite state cardinality. 
%The beam sampler combines dynamic programming with slice sampling \cite{Neal2003} in order to sample from the hierarchical Dirichlet process inherent to the iHMM.
The ``beam sampling'' approach  \cite{vanGael2008}, which we can apply largely without modification, is to truncate this sum by introducing a ``slice''  (see \cite{Neal2003}) auxiliary variable $\mathcal{U} = \{u_1, u_2, \ldots,u_T\}$ at each time step.  The auxiliary variables are chosen in such a way as to automatically limit each sum in the forward pass to a finite number of terms while still allowing all possible durations.% in the asymptotic sampling limit.

>>>>

A recent contribution to inference 
in 
%is 
the infinite Hidden Markov Model (iHMM) \cite{Beal2002} suggests a way around truncation \cite{vanGael2008}.  The iHMM is an HMM with a countable number of states.  Computing the forward message for a forward filtering, backward sampler for the latent states in an iHMM also requires a sum over a countable number of elements.  
% which can be dealt with using the beam sampler \cite{vanGael2008}. Here, instead of a %potentially infinite duration-space cardinality, the iHMM faces a potentially infinite state cardinality. 
%The beam sampler combines dynamic programming with slice sampling \cite{Neal2003} in order to sample from the hierarchical Dirichlet process inherent to the iHMM.
The ``beam sampling'' approach  \cite{vanGael2008}, which we can apply largely without modification, is to truncate this sum by introducing a ``slice'' \cite{Neal2003} auxiliary variable $\mathcal{U} = \{u_1, u_2, \ldots,u_T\}$ at each time step.  The auxiliary variables are chosen in such a way as to automatically limit each sum in the forward pass to a finite number of terms while still allowing all possible durations.% in the asymptotic sampling limit.
